Learning Fast Quadruped Robot Gaits with the RL PoWER Spline Parameterization

### 2012-06-10 paper_aimsa/abstract.tex
Abstract rewrite, many edits and suggestions for improvement.

A reinforcement learning approach which can evolve the policy parameterization dynamically during
the learning process for quadruped robot gait . By gradually increasing the representational
power of the policy parameterization, it manages to find better
policies faster than HyperNEAT, a state-of-the-art evolutionary ANN algorithm that achieved the best performance in previous studies.

Expectation-Maximization-based reinforcement learning algorithm
is used. The test was done on a physical quadrupedal robot(QuadraTot). The results show that the evolving policy parameterization combined with expectation-maximization (RL PoWER)
outperforms substantially the HyperNEAT, by having
faster convergence and higher final
reward. According to the statistics collected, the RL PoWER achieves 16.3\%
advantage over the HyperNEAT, the best method tested on QuadraTot in previous studies.

>>>>

%A reinforcement learning approach which can evolve the policy
%parameterization dynamically during the learning process for quadruped
%robot gait \edit{???}. By gradually increasing the representational
%power of the policy parameterization, it manages to find better
%policies faster than HyperNEAT, a state-of-the-art evolutionary ANN
%algorithm that achieved the best performance in previous studies.
%
%Expectation-Maximization-based reinforcement learning algorithm is
%used. The test was done on a physical quadrupedal robot
%(QuadraTot). The results show that the evolving policy
%parameterization combined with expectation-maximization (RL PoWER)
%outperforms substantially the HyperNEAT, by having faster convergence
%and higher final reward. According to the statistics collected, the RL
%PoWER achieves 16.3\% advantage over the HyperNEAT, the best method
%tested on QuadraTot in previous studies.

Legged robots are uniquely privileged over their wheeled counterparts
in their potential to access rugged terrain. However, designing
walking gaits by hand for legged robots is a difficult and
time-consuming process, so we seek algorithms for learning such gaits
automatically using real world experimentation. Numerous previous
studies have examined a variety algorithms for learning gaits, using
an assortment of different robots. It is often difficult to compare
the algorithmic results from one study to the next, because the
conditions and robots used vary. With this in mind, we have used an
open-source, 3D printed quadruped robot called QuadraTot, so the results may be
verified, and hopefully improved upon, by any group so
desiring. Because many robots do not have accurate simulators, we test
gait-learning algorithms entirely on the physical robot. Previous
studies using the QuadraTot have compared parameterized splines, the
HyperNEAT generative encoding, and (Kyrre's algorithm). Here we
compare these results to an algorithm called Policy learning by
Weighting Exploration with the Returns, or RL PoWER. We report that
this algorithm has learned the fastest gait yet reported in the
literature, 16.3\% faster \edit{different number than in results?} than reported for HyperNEAT and \edit{???
compared to Kyrre's?}. In addition, the learned gaits are less taxing
on the robot and more repeatable than previous record-breaking gaits.

### 2012-06-10 paper_aimsa/conclusion.tex
Abstract rewrite, many edits and suggestions for improvement.

We have presented a new reinforcement-learning-based algorithm for
optimizing a quadrupedal gait for linear speed. We implemented and
tested six learning strategies for parameterized gaits and compared
them to gaits produced by neural networks evolved with the HyperNEAT
generative encoding. Though both methods resulted in an improvement
over the robot’s previous \naive gaits, based on the statistics
collected, RL PoWER has a more elegant and consistent performance.

Though over 900 trials have been made to investigate the applicability
of RL PoWER to quadruped robots. It is difficult to gather the enough
trials that would be necessary to properly rank the methods. One
direction for future work could be to obtain many more trials. But due
to the physical limitations, obtaining One solution to this is
simulation. Because of the low cost of simulation, it would produce
the necessary volume of trials to allow the learning methods to be
effective, and the hardware trials would serve to continuously ground
and refine the simulator. \cite{glette2012evolution-of-locomotion-in-a-simulated}

One guess led by this study is that for feedback oriented tasks,
reinforcement learning methods are more fit in natural. Despite the
complexities of HyperNEAT, a simpler algorithm on the code level
delivered better performance in general.  Also, evolvable spline
interpolation is shown to be simple and representationally powerful at
the same time. Evolvable splines can serve as a general representation for various other
learning problems.

>>>>

We have presented results from a recently introduced reinforcement-learning-based algorithm for
optimizing a quadrupedal gait for linear speed, tested on a physical robot. \edit{Next sentence mistakenly copied?} We implemented and
tested six learning strategies for parameterized gaits and compared
them to gaits produced by neural networks evolved with the HyperNEAT
generative encoding. Though both methods resulted in an improvement
over the robot’s previous \naive gaits, based on the statistics
collected, RL PoWER has a more elegant and consistent performance.

Over 900 trials have been made to investigate the applicability
of RL PoWER to quadruped robots. It is difficult to gather the enough
trials that would be necessary to properly rank the methods. One
direction for future work could be to obtain many more trials. But due
to the physical limitations, obtaining one solution to this is
simulation. Because of the low cost of simulation, it would produce
the necessary volume of trials to allow the learning methods to be
effective, and the hardware trials would serve to continuously ground
and refine the simulator. \cite{glette2012evolution-of-locomotion-in-a-simulated} \edit{what is this citation for?}

One view supported by this study is that for feedback oriented tasks,
reinforcement learning methods are more fit and natural. Despite the
complexities of HyperNEAT, a simpler algorithm on the code level \edit{???}
delivered better performance in general.  Also, evolvable spline
interpolation is shown to be simple and representationally powerful at
the same time. Evolvable splines can serve as a general representation for various other
learning problems.\edit{probably rewrite this last paragraph}

### 2012-06-10 paper_aimsa/experimentalSetup.tex
Abstract rewrite, many edits and suggestions for improvement.

The quadruped robot used was assembled from parts purchased online and
parts printed by the Objet Connex 500 3-D Printing System. The robot
actuation system consists of 5 AX-18+ Dynamixel servos and 4 AX-12+
Dynamixel servos: one inner joint with one AX-18+ servo and one outer
joint with AX-12+servo in each of the four legs, and one AX-18+ servo
at the center. To avoid the formerly reported \edit{???} AX-18+ servos are used
in this robot because of their stronger actuation power than that of
AX-12+. Each servo could be set to a position in the range [0, 1023]
by using pydynamixel library, corresponding to a physical range [-120$^\circ$,
  +120$^\circ$]. Also, to prevent collisions with the robot body, the control
module filter out the commands to a safe range. This range was [-85$^\circ$,
  +60$^\circ$] for the inner leg servos, [-113$^\circ$, +39$^\circ$] for the outer leg
servos, and [-28$^\circ$, +28$^\circ$] for the center hip servo. In the studies of
this paper, tethered cables powered both the computer and the
servos. It measures approximately 39.5 centimeters from leg to
opposite leg in the crouch position showed in \figref{quadratotWhiteBooties}.

Our performance metric was the displacement over the evaluation period of 12 seconds
for each. Same as \cite{yosinski2011evolving-robot-gaits}, the displacement was
measured using a Wii remote that was placed on the ceiling. Different
from the original model described in \cite{yosinski2011evolving-robot-gaits}, the quadruped robot was
equipped with a three-infrared-LED cluster on top rather than just
one. The reason for this setup is that when fierce gaits were
executed, the Wii remote loses tracking of the robot position due to
the limited visible angle of a single LED.  These three LEDs were
place tightly together to act as one signal emitter. Each LED was
tilted outwards in order to maximize the visible range. A
separate tracking server ran on the robot PC interacted with the Wii
remote via bluetooth by using the CWiid library.  A corresponding
client communicates with this server via socket connection. Our
fitness evaluation code talks with the Wii remote by using this client
and updates the position from beginning to end during each run. The
metric for evaluating gaits was the Euclidian distance the robot
traveled during a 12-second run on flat surface. Previous work done in
\cite{yosinski2011evolving-robot-gaits} and \cite{clune2009evolving-coordinated-quadruped} suggested that extremely fierce gaits are not viable in
general. These gaits tend to either overburden the servos or flip the
robot. Thus, RL PoWER were tested after carefully cropping out the
extremely fierce gaits each time. As described earlier, each leg has
two joints, inner joint j1 and outer j2. The position of one leg is
determined by the sum of the values of j1 and j2. Extremely fierce
gaits usually have relatively small values of $g$. As shown in \figref{smartCrop}, this cropping method works
by mapping the wild gaits onto the boundary of a quasi-triangular area in
the two-dimensional space of j1 and j2.

\figp{smartCrop}{1}{The red line is the cropping border, $g = j1 +  j2$. As shown in \figref{sixLegPositions}}

\figp{sixLegPositions}{1}{The left two positions are allowed; middle
  two positions are on the red border line from \figref{smartCrop}; right two positions are deemed as ``extremely fierce'', thus are disallowed.}

>>>>

The quadruped robot used, QuadraTot, was assembled from parts purchased online and
parts printed by the Objet Connex 500 3-D Printing System. The robot
actuation system consists of 5 AX-18+ Dynamixel servos and 4 AX-12+
Dynamixel servos: one inner joint with one AX-18+ servo and one outer
joint with AX-12+servo in each of the four legs, and one AX-18+ servo
at the center. To avoid the formerly reported \edit{???} AX-18+ servos are used
in this robot because of their stronger actuation power than that of
AX-12+. Each servo could be set to a position in the range [0, 1023]
by using pydynamixel library, corresponding roughly to a physical range [-120$^\circ$,
  +120$^\circ$]. Also, to prevent collisions with the robot body, the control
module filter out the commands to a safe range. This range was [-85$^\circ$,
  +60$^\circ$] for the inner leg servos, [-113$^\circ$, +39$^\circ$] for the outer leg
servos, and [-28$^\circ$, +28$^\circ$] for the center hip servo. In the studies of
this paper, tethered cables powered both the computer and the
servos. It measures approximately 39.5 centimeters from leg to
opposite leg in the crouch position showed in \figref{quadratotWhiteBooties}.

Our performance metric was the displacement over the evaluation period of 12 seconds
for each. Same as \cite{yosinski2011evolving-robot-gaits}, the displacement was
measured using a Wii remote that was placed on the ceiling. Different
from the original model described in \cite{yosinski2011evolving-robot-gaits}, the quadruped robot was
equipped with a three-infrared-LED cluster on top rather than just
one. The reason for this setup is that when fierce gaits were
executed, the Wii remote loses tracking of the robot position due to
the limited visible angle of a single LED.  These three LEDs were
place tightly together to act as one signal emitter. Each LED was
tilted outwards in order to maximize the visible range. A
separate tracking server ran on the robot PC interacted with the Wii
remote via bluetooth by using the CWiid library.  A corresponding
client communicates with this server via socket connection. Our
fitness evaluation code talks with the Wii remote by using this client
and updates the position from beginning to end during each run. The
metric for evaluating gaits was the Euclidian distance the robot
traveled during a 12-second run on flat surface. Previous work done in
\cite{yosinski2011evolving-robot-gaits} and \cite{clune2009evolving-coordinated-quadruped} suggested that extremely fierce gaits are not viable in
general. These gaits tend to either overburden the servos or flip the
robot. \edit{Explain this next part on smart cropping better.} Thus, RL PoWER were tested after carefully cropping out the
extremely fierce gaits each time. As described earlier, each leg has
two joints, inner joint j1 and outer j2. The position of one leg is
determined by the sum of the values of j1 and j2. Extremely fierce
gaits usually have relatively small values of $g$. As shown in \figref{smartCrop}, this cropping method works
by mapping the wild gaits onto the boundary of a quasi-triangular area in
the two-dimensional space of j1 and j2.

\figp{smartCrop}{1}{The red line is the cropping border, $g = j1 +  j2$. As shown in \figref{sixLegPositions} \edit{???}}

\figp{sixLegPositions}{1}{The left two positions are allowed; middle
  two positions are on the red border line from \figref{smartCrop};
  right two positions are deemed as ``extremely fierce'', thus are
  disallowed.   \\\editbox{Smart crop needs to be explained somewhere, probably briefly in this caption and more extensively in the text.}}

### 2012-06-10 paper_aimsa/introduction.tex
Abstract rewrite, many edits and suggestions for improvement.

\figp{quadratotWhiteBooties}{.75}{The quadratot robot used in this research. The white booties are newly added to prevent sliding on the surface thus minimize measure error.}

Various learning algorithms have been proved to be effective for
legged robots. Algorithms such as HyperNEAT \cite{yosinski2011evolving-robot-gaits},
GA \cite{chernova2004an-evolutionary-approach-to-gait} and others \cite{hornby2005autonomous-evolution-of-dynamic} \cite{zykov2004evolving-dynamic-gaits} \cite{tellez2006evolving-the-walking-behaviour} \cite{valsalam2008modular-neuroevolution-for-multilegged} have been tested to be effective for automatic learning gaits for robots. Despite competetive performance, a major
task that are usually hidden in the published results lies in tuning the parameters for
these evolutionary algorithms \cite{kormushev2011bipedal-walking-energy}. Here we present a
different way for learning gaits using reinforcement learning and expectation maximization, called RL PoWER. In our
experiment, the main focus of the research is on the applicability of RL PoWER to quadruped robot gait learning. Another motivation is to compare the state-of-the-art neural network algorithm, HyperNEAT, with our proposed method in quadruped robot gait learning.

@@

As in \cite{yosinski2011evolving-robot-gaits}, the gait learning problem is defined to find a
gait that maximizes some specific metric. Mathematically, we define a
gait as a function that specifies a vector of commanded motor
positions for a robot over time. Gaits without feedback---also
called open-loop gaits---can be defined as

>>>>

\figp{quadratotWhiteBooties}{.75}{The open source, 3D printed
  QuadraTot robot used in this research. The white printed booties are
  new addition to prevent sliding on surfaces and to minimize measurement
  error.}

Various learning algorithms have been proved to be effective for
legged robots. Algorithms such as HyperNEAT
\citep{yosinski2011evolving-robot-gaits}, Genetic Algorithms
\citep{chernova2004an-evolutionary-approach-to-gait} and others
\citep{hornby2005autonomous-evolution-of-dynamic,
  zykov2004evolving-dynamic-gaits,
  tellez2006evolving-the-walking-behaviour,
  valsalam2008modular-neuroevolution-for-multilegged} have been tested
to be effective for automatic learning gaits for robots. Despite
competitive performance, a major task is usually hidden in the
published results: tuning the parameters for these evolutionary
algorithms \citep{kormushev2011bipedal-walking-energy}. Here we
present results using a different way of learning gaits: a
Reinforcement Learning algorithm called Policy learning by Weighting
Exploration with the Returns, or RL PoWER, proposed by
\cite{kober2009learning-motor-primitives}. In our experiment, the main
focus of the research is on the applicability of RL PoWER to quadruped
robot gait learning. Another motivation is to compare the
state-of-the-art neural network algorithm, HyperNEAT, with our
proposed method in quadruped robot gait learning.

@@

As in \cite{yosinski2011evolving-robot-gaits}, we define the gait learning problem to be the search for a
gait that maximizes some specific metric. Mathematically, we define a
gait as a function that specifies a vector of commanded motor
positions for a robot over time. Gaits without feedback---also
called open-loop gaits---can be defined as

### 2012-06-10 paper_aimsa/main.tex
Abstract rewrite, many edits and suggestions for improvement.

\title{RL-PoWER, Evolving Policy Parameterization with Reinforcement Learning and its Application to Quadrupedal Walking Optimization}
\author{Haocheng Shen, Jason Yosinski, Petar Kormushev, \and Hod Lipson  \\
  \mbox{}\\
Cornell University, 239 Upson Hall, Ithaca, NY 14853 \\
\texttt{hs454@cornell.edu, yosinski@cs.cornell.edu}}

>>>>

%\title{RL-PoWER, Evolving Policy Parameterization with Reinforcement Learning and its Application to Quadrupedal Walking Optimization}
\title{Learning Fast Quadruped Robot Gaits with the RL PoWER Spline Parameterization}
\author{Haocheng Shen, Jason Yosinski, Petar Kormushev, \and Hod Lipson  \\
  \mbox{}\\
Cornell University, 239 Upson Hall, Ithaca, NY 14853 \\
\texttt{hs454@cornell.edu,yosinski@cs.cornell.edu}}

### 2012-06-10 paper_aimsa/methods.tex
Abstract rewrite, many edits and suggestions for improvement.

Here we used an RL approach to change the complexity of the policy
representation dynamically while the trial is running. In
\cite{kormushev2011bipedal-walking-energy}'s studies on reducing energy
consumption for bipedal robots, a mechanism that can
evolve the policy parameterization was used. The method starts from a
very simple parameterization and gradually increases its
representational capability. The method was tested to be capable of generating
an adaptive policy parameterization that can accommodate increasingly
more complex policies. Presented in the studies of \cite{kormushev2011bipedal-walking-energy}, the policy
generated by this approach can reach the global optimum at a fast
rate when applied to the energy reduction problem. Another property found about this method is its chance of converging to a suboptimal solution is reduced, because in the lower-dimensional representation this effect is less exhibited.

@@

\cite{kober2009learning-motor-primitives} proposed a RL algorithm
called Policy learning by Weighting Exploration with the
Returns(PoWER), which is based on Expectation-Maximization algorithm
(EM). The proposed technique for evolving the policy parameterization
is a combination with this EM-based RL algorithm, named PoWER \cite{kober2009learning-motor-primitives}. The reason for using this is its relatively fewer parameters that need tuning. We
evolved the policy parameterization only on those past trials ranked
the highest by the importance sampling technique used by the PoWER
algorithm. The intuition behind is that highly ranked
parameterizations have more potential to evolve even better in the
future. Besides, evolving all the parameterizations increases the exploring space. Since our experiment
is done on a physical robot, explore all the variations of every
parameterization is not practical. Future work may incorporate simulations into the studies, as illustrated in \cite{bongard2006resilient-machines-through}.

For the experiment, we set 3 knots for each servo and there are 8
servos in total. The servo in the hip is not used in
our experiment. Previous work has verified that quadruped gaits
perform better when they are coordinated \cite{clune2009evolving-coordinated-quadruped, clune2011on-the-performance-of-indirect-encoding, valsalam2008modular-neuroevolution-for-multilegged}. For each spline, we calculate its corresponding parameterized gait for one unit time cycle. Given that, then apply the same
pattern to every cycle throughout the 12 seconds of one
trial. Specifically, each spline (a set of 3 knots) is interpreted to its corresponding servo positions as shown in \eqnref{rl} and \tabref{params}.

>>>>

Here we used an RL approach to change the complexity of the policy
representation dynamically while the trial is running. In
earlier studies on reducing energy
consumption for bipedal robots \citep{kormushev2011bipedal-walking-energy}, a mechanism that can
evolve the policy parameterization was used. The method starts from a
very simple parameterization and gradually increases its
representational capability. The method was tested to be capable of generating
an adaptive policy parameterization that can accommodate increasingly
more complex policies. Presented in the studies of \cite{kormushev2011bipedal-walking-energy}, the policy
generated by this approach can reach the global optimum at a fast
rate when applied to the energy reduction problem. Another property found about this method is its chance of converging to a suboptimal solution is reduced, because in the lower-dimensional representation this effect is less exhibited.

@@

\cite{kober2009learning-motor-primitives} proposed a RL algorithm
called Policy learning by Weighting Exploration with the
Returns (RL PoWER), which is based on Expectation-Maximization algorithm
(EM). The proposed technique for evolving the policy parameterization
is a combination with this EM-based RL algorithm, named PoWER \cite{kober2009learning-motor-primitives}. \edit{Check the prev sentence... I don't get it} The reason for using this is its relatively fewer parameters that need tuning. We
evolved the policy parameterization only on those past trials ranked
the highest by the importance sampling technique used by the PoWER
algorithm. The intuition behind is that highly ranked
parameterizations have more potential to evolve even better in the
future. Besides, evolving all the parameterizations increases the exploring space. Since our experiment
is done on a physical robot, exploring all the variations of every
parameterization is not practical. Future work may incorporate simulations into the studies, as illustrated in \cite{bongard2006resilient-machines-through}.

For the experiment, we set the splines to have 3 knots for each servo, and there are 8
servos in total. The servo in the hip is not used in
our experiment. Previous work has verified that quadruped gaits
perform better when they are coordinated \citep{clune2009evolving-coordinated-quadruped, clune2011on-the-performance-of-indirect-encoding, valsalam2008modular-neuroevolution-for-multilegged}. For each spline, we calculate its corresponding parameterized gait for one unit time cycle \edit{how long is this?} and then apply the same
pattern to every cycle throughout the 12 seconds of one
trial. Specifically, each spline (a set of 3 knots) is interpreted to its corresponding servo positions as shown in \eqnref{rl} and \tabref{params}.

### 2012-06-10 paper_aimsa/results.tex
Abstract rewrite, many edits and suggestions for improvement.

\begin{table}
\begin{center}
\begin{tabular}{|r|c|c|c||c|}
\hline
                                         & Average & Std. Dev \\
\hline                                    
\hline                                    
RL PoWER                       & 7.62  &    2.1   \\
\hline
HyperNEAT                      & 6.28   &   1.26   \\
\hline
\end{tabular}
\caption{The average and standard deviation of the best gaits found
  for both algorithm during each of three runs, in cm/sec.}  \tablabel{results}
\end{center}
\end{table}

We believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
The single best gait found during this study had a speed of 11.09 centimeters/sec,  14.7\% better than the best HyperNEAT gait. \figref{rlPattern} shows a typical RL PoWER gait that had high fitness. Compared with \figref{neat_110115_211410_00000_002_filt_zoom}, the pattern of RL PoWER's motion is less
complex but more regular than the HyperNEAT, but producing higher speed. As shown in \figref{rlPattern}
, multiple motors are more coordinated than the HyperNEAT gait. 
\figp{rlPattern}{0.75}{The motor positions of a typical well performing gait over one cycle. It matches exactly the shape of its corresponding spline representation. This pattern shows the simplicity and power of geometric splines representations.}

\figp{neat_110115_211410_00000_002_filt_zoom}{0.75}{A typical well performing HyperNEAT gait's pattern of motor positions}

>>>>

\begin{table}
\begin{center}
\begin{tabular}{|r|c|c|c||c|}
\hline
                                         & Average & Std. Dev \\
\hline                                    
\hline                                    
RL PoWER                       & {\bf 7.62}  &    2.1   \\
\hline
HyperNEAT                      & 6.28   &   1.26   \\
\hline
\end{tabular}
\caption{The average and standard deviation of the best gaits found
  for both algorithm during each of three runs, in cm/sec. \edit{Add the best gait each algorithm found as well.}}
\tablabel{results}
\end{center}
\end{table}

We believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
The single best gait found during this study had a speed of 11.09 centimeters/sec,  14.7\% better \edit{different number than in abstract?} than the best HyperNEAT gait. \figref{rlPattern} shows a typical RL PoWER gait that had high fitness. Compared with \figref{neat_110115_211410_00000_002_filt_zoom}, the pattern of RL PoWER's motion is less
complex but more regular than the HyperNEAT, but producing higher speed. As shown in \figref{rlPattern}
, multiple motors are more coordinated than the HyperNEAT gait.

\figp{rlPattern}{0.75}{The motor positions of a typical well
  performing gait over one cycle. It matches exactly the shape of its
  corresponding spline representation \edit{what does this sentence mean?}. This pattern shows the
  simplicity and power of geometric splines representations \edit{how does it show this?}.}

\figp{neat_110115_211410_00000_002_filt_zoom}{0.75}{A typical well
  performing HyperNEAT gait's pattern of motor positions, reproduced
  from \cite{yosinski2011evolving-robot-gaits}.}
